Autopolyploid establishment through polygenic adaptation

### 2025-01-22 doc/paper/paper.tex
first draft, editing

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
In particular, polyploids are thought to establish mainly in novel, unoccupied
habitats where they evade MCE (for instance at range edges, or after local
extinction due to environmental change) \citep{griswold2021}.
However, such peripheral habitats are likely to present adaptive challenges to
establishment \citep{kawecki2008adaptation}, and if polyploids are able to
colonize such habitats at an appreciable rate, they must somehow be better
adapted to local conditions, or more able to adapt to those conditions despite
inbreeding and maladaptive migration, compared to diploids.
Indeed, when the source population is dominated by diploids, the probability
that a migrant individual is tetraploid will be very small ($O(u^2)$ if $u$ is
the probability that a diploid in the source population produces an unreduced
gamete in meiosis), so that tetraploids need a substantial advantage during
colonization if they are to establish before diploids do.

@@

Here we develop a rather general model for the establishment of a mixed-ploidy
population in a novel, unoccupied habitat based on \cite{barton2018}.
In order to establish in the novel habitat, the population has to adapt to
local environmental conditions.
We assume fitness is determined by directional selection on a single polygenic
trait which can be interpreted as log fitness at low density in the new
habitat.
As in \cite{barton2018}, we assume the trait follows the infinitesimal
model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
infinitesimal model \citep{turelli2017}).
We extend the infinitesimal model, and the approach for exact simulation of
trait evolution under the infinitesimal model, to mixed-ploidy populations.
This is a first contribution of the present paper.
We then use simulations to study tetraploid establishment, both from single
migrants and under continuous migration from a predominantly diploid source
population, examining the effects of autopolyploid genetics,
maladaptive migration, selfing and assortative mating on the probability that
autotetraploids establish in the novel habitat.

@@

\begin{table}[t]
\caption{Glossary of the notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N$ & total population size\\
$N_k$ & population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & rate of self-fertilization in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

An alternative, and for our purposes useful, way to characterize the model
is to write $Z_{ij} = Y_i + Y_j$, where $Y_i$ and $Y_j$
are independent Gaussian random variables $Y_i \sim \Normal\left(\frac{z_i}{2},
V_i\right)$ (and similarly for $Y_j$).
We refer to $Y_i$ as the (random) \textit{gametic value} of individual $i$, and
to $V_i$ as the \textit{gametic segregation variance} of individual $i$.
This formulation is helpful in that it highlights that Mendelian segregation
occurs independently in both parents when gametes are produced, which then
combine additively to determine the offspring trait value.
This model applies readily to an autopolyploid population expressing a trait
with infinitesimal genetics.
For instance, when an autotetraploid produces a diploid gamete, it will pass on
half its trait value to the gamete on average, with a variance determined by
the details of tetraploid meiosis (which are not, in general, identical to
those of diploid meiosis, see below).

In a finite population, the segregation variance will decay over time
as the population becomes more inbred.
Indeed, Mendelian segregation generates less variation among the gametes
produced by individual $i$ when that individual is more inbred, as segregation
at homozygous loci does not generate any variation.
When $F_i$ is the inbreeding coefficient relative to some ancestral reference
population with gametic segregation variance $V$ (i.e. the probability that two
genes at a locus in individual $i$ sampled without replacement are identical by
descent), the gametic segregation variance of individual $i$ will be
$V_i = (1-F_i)V$.
This holds for both diploids and tetraploids (\cref{s-sec:tetinbred}, also
\cite{moody1993}).

@@

If we would somewhat naively assume that the allelic effects underlying an
additive quantitative trait are identical across ploidy levels, 
a tetraploid offspring from a cross between two diploids would yield, on average,
a trait value which is the sum of the parental trait values, with a
distribution that depends on the variance generated by the process producing
unreduced gametes.
This is not likely to reflect biological reality: tetraploids do not tend to
have, for instance, twice the size of their diploid progenitors on average.
Furthermore, the genetic variance at Hardy-Weinberg and linkage equilibrium
(HWLE) in a large non-inbred tetraploid population will be twice that of their
diploid counterparts under such assumptions, which is similarly unrealistic
\citep{clo2022}.

@@

For the production of diploid gametes by a $k$-ploid individual, the
segregation variance depends not only on the segregation variance in the base
population ($V$) and the inbreeding coefficient ($F$), but also on the detailed
assumptions on how these aberrant meiotic processes occur.
Importantly however, these details only affect the gametic segregation variance
through the quantity $\alpha_k$, which is the probability that a $k$-ploid
transmits two copies of the same homolog to a diploid gamete.
Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
individual carries two copies of the same homolog, is the probability of
\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
1/6 \citep{stift2008}.
The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
formation through so-called \textit{first} and \textit{second division
restitution} \citep{bretagnolle1995,storme2013}.
We summarize the expressions for the gametic segregation variance in
\cref{tbl:segvar}.
Detailed derivations can be found in \cref{s-sec:segvar}.

@@

Dominance and inbreeding depression may strongly affect the complicated
relationship between selfing, migration load and tetraploid establishment
probability.
\cite{griswold2021} performed a study somewhat similar to ours, but did include
a simple model of inbreeding depression.
The author focused on the case where local fitness is determined by a single
biallelic locus, and investigated the interaction between inbreeding depression
and migration load (where inbreeding depression is modeled as an immediate
fitness reduction in offspring produced by selfing).
In his model, inbreeding depression is different between cytotypes, so that
tetraploids are able to produce more offspring through selfing relative to
diploids, who have to rely more on outcrossing.
However, outcrossing incurs maladaptive gene flow, and thereby puts the
diploids at a disadvantage. 
He found that autotetraploids can establish when adaptation in the
peripheral habitat is conferred by recessive alleles (so that migration load is
expressed when migrant alleles are rare) and when inbreeding depression in
tetraploids is lower than in diploids.
It would be very interesting to combine the infinitesimal framework with some
form of inbreeding depression to investigate in a somewhat more realistic model
whether the combination of maladaptive migration and differential inbreeding
depression could explain the prevalence of polyploid subpopulations at range
edges.

>>>>

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
In particular, polyploids are thought to establish mainly in novel, unoccupied
habitats where they evade MCE (for instance at range edges, or after local
extinction due to environmental change) \citep{griswold2021}.
However, such peripheral habitats are likely to present adaptive challenges to
establishment \citep{kawecki2008adaptation}, and if polyploids are able to
colonize such habitats at an appreciable rate, they must somehow be better
adapted to local conditions, or more able to adapt to those conditions despite
inbreeding and maladaptive migration, compared to diploids.
Indeed, when the source population is dominated by diploids, the probability
that a migrant individual is tetraploid will be very small ($\sim u^2$, where
$u$ is the probability that a diploid produces an unreduced gamete in meiosis),
so that tetraploids need a substantial advantage during colonization if they
are to establish before diploids do.

@@

Here we develop a model for the establishment of a mixed-ploidy population in a
novel, unoccupied habitat based on \cite{barton2018}.
In order to establish in the novel habitat, the population has to adapt to
local environmental conditions.
We assume fitness is determined by directional selection on a single polygenic
trait, which can be interpreted as log fitness at low density in the new
habitat.
As in \cite{barton2018}, we assume the trait follows the infinitesimal
model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
infinitesimal model \citep{turelli2017}).
We extend the infinitesimal model, and the approach for exact simulation of
trait evolution under the infinitesimal model, to mixed-ploidy populations.
This is a first contribution of the present paper.
We then use simulations to study tetraploid establishment, both from single
migrants and under continuous migration from a predominantly diploid source
population, examining the effects of autopolyploid genetics,
maladaptive migration, selfing and assortative mating on the probability that
autotetraploids establish in the novel habitat.

@@

\begin{table}[t]
\caption{Glossary of the notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N$ & total population size\\
$N_k$ & population size of the $k$-ploid cytotype \\
$\pi_k$ & deterministic equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & rate of self-fertilization in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

An alternative, and for our purposes useful, way to characterize the model
is to write $Z_{ij} = Y_i + Y_j$, where $Y_i$ and $Y_j$
are independent Gaussian random variables $Y_i \sim \Normal\left(\frac{z_i}{2},
V_i\right)$ (and similarly for $Y_j$).
We refer to $Y_i$ as the (random) \textit{gametic value} of individual $i$, and
to $V_i$ as the \textit{gametic segregation variance} of individual $i$.
This formulation is helpful in that it highlights that Mendelian segregation
occurs independently in both parents to produce gametes, which then combine
additively to determine the offspring trait value.
This model applies readily to an autopolyploid population expressing a trait
with infinitesimal genetics.
However, the segregation variance will be determined by the details of
tetraploid meiosis, which differ from those of diploid meiosis (see below).

In a finite population, the segregation variance will decay over time as the
population becomes more inbred (Mendelian segregation at homozygous loci does
not generate any variation).
When $F_i$ is the inbreeding coefficient relative to some ancestral reference
population with gametic segregation variance $V$ (i.e. the probability that two
genes at a locus in individual $i$ sampled without replacement are identical by
descent), the gametic segregation variance of individual $i$ will be
$V_i = (1-F_i)V$.
This holds for both diploids and tetraploids (\cref{s-sec:tetinbred}, also
\cite{moody1993}).

@@

If we would naively assume that the allelic effects underlying an additive
trait are identical across ploidy levels, 
a tetraploid offspring from a cross between two diploids would have, on average,
a trait value which is the sum of the parental trait values.
This is not likely to reflect biological reality: tetraploids do not tend to
have, for instance, twice the size of their diploid progenitors on average.
Furthermore, the genetic variance at Hardy-Weinberg and linkage equilibrium
(HWLE) in a large non-inbred tetraploid population will be twice that of their
diploid counterparts under such assumptions, which is similarly unrealistic
\citep{clo2022}.

@@

The gametic segregation variance associated with the production of diploid
gametes depends not only on the segregation variance in the base population
($V$) and the inbreeding coefficient ($F$), but also on the detailed
assumptions of how the meiotic process takes place.
Importantly however, the latter only affect the gametic segregation variance
through the quantity $\alpha_k$, which is the probability that a $k$-ploid
transmits two copies of the same homolog to a diploid gamete.
Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
individual carries two copies of the same homolog, is the probability of
\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
1/6 \citep{stift2008}.
The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
formation through so-called \textit{first} and \textit{second division
restitution} \citep{bretagnolle1995,storme2013}.
We summarize the expressions for the gametic segregation variance in
\cref{tbl:segvar}.
Detailed derivations can be found in \cref{s-sec:segvar}.

@@

Dominance and inbreeding depression may strongly affect the complicated
relationship between selfing, migration load and tetraploid establishment
probability.
\cite{griswold2021} performed a study somewhat similar to ours, but did include
a simple model of inbreeding depression.
The author focused on the case where local fitness is determined by a single
biallelic locus, and investigated the interaction between inbreeding depression
and migration load (where inbreeding depression is modeled as a fixed fitness
reduction in offspring produced by selfing).
In his model, inbreeding depression is different between cytotypes, so that
tetraploids are able to produce more offspring through selfing relative to
diploids, who have to rely more on outcrossing.
However, outcrossing incurs maladaptive gene flow, and thereby puts the
diploids at a disadvantage. 
He found that autotetraploids can establish when adaptation in the
peripheral habitat is conferred by recessive alleles (so that migration load is
expressed when migrant alleles are rare) and when inbreeding depression in
tetraploids is lower than in diploids.
It would be very interesting to combine the infinitesimal framework with some
form of inbreeding depression to investigate in a somewhat more realistic model
whether the combination of maladaptive migration and differential inbreeding
depression could explain the prevalence of polyploid subpopulations at range
edges.
